---
title: "hw4_yw3204"
author: "wyh"
date: "10/13/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.
```{r}
library(plyr)
debt <- read.csv("debt.csv", as.is = TRUE)
dim(debt)
head(debt)

# a 
# calculate the mean of the growth column of a data frame
mean.growth <- function(df) {
  return (mean(df$growth))
}

# b
avg_g <- daply(debt, 'Country', mean.growth)
signif(avg_g, 3) # print
```

## 2.
```{r}
avg_g1 <- daply(debt, 'Year', mean.growth)
signif(avg_g1, 3) # print

# plot
plot(c(1946:2009), avg_g1, xlab = 'year', ylab = 'gdp_growth')
```

## 3.
```{r}
# a
c1 <- cor(debt$growth, debt$ratio)
signif(c1, 3)

# b
# calculate the correlation between growth and ratio column in a data frame
cor_fun <- function(df) {
  return (cor(df$growth, df$ratio))
}

c2 <- daply(debt, 'Country', cor_fun)
#mean(c2) #check, if you may
hist(c2, breaks = 10, main = "cor_hist_ctry", xlab = "cor")

# c
c3 <- daply(debt, 'Year', cor_fun)
#mean(c3) #check
hist(c3, breaks = 10, main = "cor_hist_year", xlab = "cor")

# d
```
When splitting by country, we find that most country have negative correlation except Australia, Canada, New Zealand, Norway and Spain.
When splitting by year, we also find most year have negative correlation but for year 1978, it has the highest positive corrrelation.

## 4.
```{r}
l1 <- lm(growth ~ ratio, debt)
signif(l1$coefficients, 3)
plot(debt$ratio, debt$growth, xlab = "ratio", ylab = "growth")
abline(l1, col = "red")
```

## 5.
```{r}
ctrs <- names(c2[c2 < -0.5])
par(mfrow=c(2,2))
for(i in ctrs) {
  plot(debt[debt$Country == i, ]$ratio, debt[debt$Country == i, ]$growth, 
       xlab = "ratio", ylab = "growth", main = i)
}
```

## 6.
```{r}
# a
fra <- debt[debt$Country=="France", ]
# check
# dim(fra)

# b
n <- nrow(fra)
ng <- c()
for(i in c(1:n-1)) {
  ng[i] <- ifelse(fra$Year[i+1] == fra$Year[i]+1, fra$growth[i+1], NA)
}
ng[n] <- NA
fra$next.growth <- ng

signif(fra[fra$Year==1971, ]$next.growth, 3)
signif(fra[fra$Year==1972, ]$next.growth, 3)
```

## 7.
```{r}
add_ng <- function(df) {
  n <- nrow(df)
  ng <- c()
  for(i in c(1:n-1)) {
    ng[i] <- ifelse(df$Year[i+1] == df$Year[i]+1, df$growth[i+1], NA)
  }
  ng[n] <- NA
  df$next.growth <- ng
  return(df)
}

debt <- ddply(debt, "Country", add_ng)

ind <- debt$Country=="France" & debt$Year == 2009
signif(debt[ind, ]$next.growth, 3)
```

## 8.
```{r}
l2 <- lm(next.growth~ratio, debt)
plot(debt$ratio, debt$next.growth, xlab = "ratio", ylab = "next_growth")
abline(l2, col = 'red')
signif(l2$coefficients, 3)
```
Comparatively speaking, these two perform similarly, both of which have a slightly small negative slope coefficient.

## 9.
```{r}
l3 <- lm(next.growth~growth, debt)
plot(debt$growth, debt$next.growth, xlab = "cur_growth", ylab = "next_growth")
abline(l3, col = 'red')
signif(l3$coefficients, 3)
```
Comparatively speaking, current growth seems to be a better predictor since it has a larger coefficients.










