---
title: "hw2"
author: "NAME: Yuhao Wang, UNI: yw3204"
date: "9/22/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part1
### i.
```{r}
# load data
housing <- read.csv("NYChousing.csv")
```

### ii.
```{r}
# no. of rows and colums
dim(housing)
```

### iii.
```{r}
apply(is.na(housing), 2, sum)
```
It is finding the number of all "NA"" records for each variable or column.

### iv.
```{r}
# remove the rows for which the variable Value is NA.
ind <- is.na(housing$Value)
ind = !ind
housing <- housing[ind, ]
```

### v.
It removes 52 rows and agrees with the results from (iii).

### vi.
```{r}
# create logarithm of "Value"
housing$logValue <- log(housing$Value, 10)
summary(housing$logValue)
```
The minimum, median, mean, and maximum values of logValue are listed above in the summary.

### vii.
```{r}
# create logarithm of "UnitCount"
housing$logUnits <- log(housing$UnitCount, 10)
```

### viii.
```{r}
# create a new variable "after1950"
housing$after1950 <- ifelse(housing$YearBuilt >= 1950, TRUE, FALSE)
```



## Part2
### i.
```{r}
# plot logValue against logUnits
library(ggplot2)
ggplot(housing, aes(x = logUnits, y = logValue)) + geom_point(aes(color = logValue)) + 
  labs(x = "logUnits", y = "logValue", title = "logValue VS logUnits")
```

### ii.
```{r}
# color data by "after1950"

# solution 1
ggplot(housing, aes(x = logUnits, y = logValue)) + geom_point(aes(color = after1950)) + 
  labs(x = "logUnits", y = "logValue", title = "logValue VS logUnits")

# solution 2
# plot(housing$logUnits, housing$logValue, col = factor(housing$after1950),
#   xlab = "logUnits", ylab = "logValue")
# legend("bottomright", legend = levels(factor(housing$after1950)), 
#   fill = unique(factor(housing$after1950)))
```
Basically, the larger the logUnits, the larger the logValue. And in general, logUnits and logValue after 1950 tend to be larger than that before 1950.


### iii.
```{r}
# calculate correlation
cor(housing[, c("logValue", "logUnits")])

ind1 <- housing$Borough == "Manhattan"
cor(housing[ind1, c("logValue", "logUnits")])

ind2 <- housing$Borough == "Brooklyn"
cor(housing[ind2, c("logValue", "logUnits")])

ind3 <- housing$after1950 == TRUE
cor(housing[ind3, c("logValue", "logUnits")])

ind4 <- housing$after1950 == FALSE
cor(housing[ind4, c("logValue", "logUnits")])
```

### iv.
```{r}
# plot logValue against logUnits for Manhattan and Brooklyn
ind5 <- housing$Borough == "Manhattan" | housing$Borough == "Brooklyn"
sub <- housing[ind5, c("logValue", "logUnits", "Borough")]
ggplot(sub, aes(x = logUnits, y = logValue)) + geom_point(aes(color = Borough)) + 
  labs(x = "logUnits", y = "logValue", title = "logValue VS logUnits in Manhattan and Brooklyn")
```

### v.
```{r}
# find the median of "Value" in Manhattan
median(housing[housing$Borough == "Manhattan", ]$Value, na.rm = TRUE)
```

### vi.
```{r}
# box plots comparing logValue across the five boroughs.
ggplot(housing, aes(x = Borough, y = logValue)) + geom_boxplot(aes(color = Borough))
```

### vii.
```{r}
# calculate median for Value for five boroughs
temp <- c()
for(name in c("Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island")) {
  temp <- c(temp, median(housing[housing$Borough == name, ]$Value, na.rm = TRUE))
}
temp
```
As above, the median of Value for Bronx, Brooklyn, Manhattan, Queens and Staten Island are 1192950, 417610, 1172362, 3611700 and 2654100.














