---
title: "Homework 6 Solutions"
output: pdf_document
---


```{r}
set.seed(0)
```

#Problem 1: Inverse Transform Method

     
Consider the *Cauchy random variable* $X$ with probability density function 
\[
f_X(x)=\frac{1}{\pi}\frac{1}{(1+x^2)}, \ \ \  -\infty <x < \infty.  
\]


(1) [5 points] Let $U$ be a uniform random variable over [0,1].  Find a transformation of $U$ that allows you to simulate $X$ from $U$.     

**Solution:**

First note that 
\[
\int_{-\infty}^x \frac{1}{\pi}\frac{1}{(1+t^2)}dt = \cdots = \frac{1}{\pi}\text{arctan}(x)+\frac{1}{2}. 
\]
Then
\[
\frac{1}{\pi}\text{arctan}(X)+\frac{1}{2} = U \longrightarrow X = \tan{(\pi(U-1/2))}
\]

(2) [5 points] Write a **R** function called **cauchy.sim** that generates $n$ simulated Cauchy random variables.  The function should have the single input **n** and should use the inverse-transformation from Part 1.  Test your function using 10 draws.

```{r}
cauchy.sim <-function(n) {
  
  U <- runif(n)
  return(tan(pi*(U-1/2)))
  
}
cauchy.sim(10)
```


(3) [5 points] Using your function **cauchy.sim**,  simulate 1000 random draws from a Cauchy distribution.  Store the 1000 draws in the vector **cauchy.draws**.  Construct a histogram of the simulated Cauchy random variable with $f_X(x)$ overlaid on the graph.   **Note:** when plotting the density curve over the histogram, include the argument  **prob = T**. 


```{r}
cauchy.draws <- cauchy.sim(1000)
hist(cauchy.draws,prob = T,main="Cauchy Draws",xlab="X",breaks=10000,xlim=c(-10,10),ylim=c(0,.4))
x <- seq(-10,10,by=.1)
lines(x,1/(pi*(1+x^2)),col="purple")
```


#Problem 2: Reject-Accept Method

Let random variable $X$ denote the temperature at which a certain chemical reaction takes place.  Suppose that $X$ has probability density function 
$$
f(x)=\begin{cases} \frac{1}{9}(4-x^2) \ \ \ \ -1 \leq x \leq 2 \\  0 \ \ \ \   \text{otherwise}
\end{cases}
$$  

Perform the following tasks:

(4) [5 points] Write a function `f` that takes as input a vector `x` and returns a vector of `f(x)` values. Plot the function between $-3$ and $3$.  Make sure your plot is labeled appropriately.

```{r}
f <- function(x) {
  ifelse((x < -1 | x > 2), 0, (1/9)*(4 - x^2))
}
x <- seq(-3, 3, by = .01)
plot(x,f(x),main="Target Distribution",ylab="f(x)",type="l",col="purple")
```

(5) [5 points] Determine the maximum of $f(x)$ and find an envelope function $e(x)$ by using a uniform density for $g(x)$ and setting $\alpha=1/\max_x\{ f(x)\}$ as in class.  Write a function `e` which takes as input a vector `x` and returns a vector of `e(x)` values.

Note that
$$
f'(x)=\begin{cases} -\frac{2x}{9} \ \ \ \ -1 \leq x \leq 2 \\  0 \ \ \ \   \text{otherwise}
\end{cases}
$$  
Setting the above equal to $0$ we find that the maximum occurs at $x = 0$ and $f(0) = 4/9$.
```{r}
f.max <- 4/9
e <- function(x) {
  ifelse((x < -1 | x > 2), Inf, f.max)
}
plot(x,f(x),main="Target Distribution",ylab="f(x)",type="l",col="purple")
lines(x,e(x),col="red")
```

(6) [5 points] Using the *Accept-Reject Algorithm*, write a program that simulates 1000 draws from the probability density function $f(x)$ from Equation \ref{e:temp}.   
```{r}
n.samps <- 1000   # Samples desired
n       <- 0      # counter for number samples accepted
samps   <- numeric(n.samps) # initialize the vector of output
while (n < n.samps) {
  y <- runif(1, min = -1, max = 2)    # random draw from g
  u <- runif(1)
  if (u < f(y)/e(y)) {
    n        <- n + 1
    samps[n] <- y
  }
}
```

(7) [5 points] Plot a histogram of your simulated data with the density function $f$ overlaid in the graph.  Label your plot appropriately. 

```{r}
hist(samps, prob = T, ylab = "f(x)",breaks=20, xlab = "x", main = "Histogram of Reject-Accept Method Draws",xlim=c(-3,3))
lines(x, f(x), lty = 2,col="purple")
```




**Problem 3: Reject-Accept Method Continued**

Consider the standard normal distribution $X$ with probability density function 
\[
f(x)=\frac{1}{\sqrt{2\pi}}\exp{\big{(}-\frac{1}{2}x^2\big{)}}, \ \ \  -\infty < x < \infty.  
\]
In this exercise, we will write a function named **normal.sim** that simulates a standard normal random variable using the **Accept-Reject Algorithm**.  

Perform the following tasks: 


(8) [5 points] Write a function **f** that takes as input a vector **x** and returns a vector of **f(x)** values. Plot the function between $-5$ and $5$.  Make sure your plot is labeled appropriately.

```{r}
f <- function(x) {
  
  return(1/sqrt(2*pi)*exp(-x^2/2))
 
}
x <- seq(-5,5,by=.01)
plot(x,f(x),main="Standard Normal Density",ylab="f(x)",type="l")
```


(9) [5 points] Let the known density $g$ be the Cauchy density defined by pdf 
\[
g(x)=\frac{1}{\pi}\frac{1}{(1+x^2)}, \ \ \  -\infty <x < \infty.  
\]
Write a function **e** that takes as input a vector **x** and constant **alpha** ($0<\alpha<1$) and returns a vector of **e(x)** values.  The envelope function should be defined as $e(x)=g(x)/\alpha$. 

```{r}
e <- function(x,alpha=1) {
  
  return(1/(alpha*pi*(1+x^2)))
 
}
```

(10) [5 points] Determine a ``good" value of $\alpha$.  To show your solution, plot both $f(x)$ and $e(x)$ on the interval $[-10,10]$.   

```{r}
x <- seq(-10,10,by=.01)
plot(x,f(x),main="Cauchy Envelope Over5Standard Normal",ylab="",type="l",ylim=c(0,.60))
lines(x,e(x,alpha=.60),main="Standard Normal Density",ylab="f(x)",col="red")
all(f(x)<e(x,alpha=.65))
```

A ``good" value of $\alpha$ is somewhere near .6.  


(11) [5 points] Write a function named **normal.sim** that simulates **n** standard normal random variables using the **Accept-Reject Algorithm**.  The function should also use the **Inverse-Transformation** from Part 1.  Test your function using **n=10** draws.  
```{r}
normal.sim <- function(n.samps) {
  n       <- 0      # counter for number samples accepted
  samps   <- numeric(n.samps) # initialize the vector of output
  while (n < n.samps) {
    y <- cauchy.sim(1) # random draw from g
    u <- runif(1)
    if (u < f(y)/e(y)) {
      n        <- n + 1
      samps[n] <- y
    }
  }
  return(samps)
}
normal.sim(10)
```

(12) [5 points] Using your function **normal.sim**,  simulate 10,000 random draws from a standard normal distribution.  Store the 10,000 draws in the vector **normal.draws**.  Construct a histogram of the simulated standard normal random variable with $f(x)$ overlaid on the graph.   **Note:** when plotting the density curve over the histogram, include the argument  **prob = T**. 

```{r}
normal.draws <- normal.sim(10000)
hist(normal.draws, prob = T, ylab = "f(x)",breaks=50, xlab = "x", main = "Histogram of Reject-Accept Method Draws")
lines(x, f(x), lty = 2)
```



# Part 3: Simulation with Built-in `R` Functions

Consider the following "random walk" procedure:


a. Start with $x=5$
b. Draw a random number $r$ uniformly between $-2$ and $1$.
c. Replace $x$ with $x+r$
d. Stop if $x \leq 0$
e. Else repeat

Perform the following tasks: 

(13) [5 points] Write a `while()` loop to implement this procedure. Importantly, save all the positive values of $x$ that were visited in this procedure in a vector called `x.vals`, and display its entries.
```{r}
x.vals <- 5
i <- 1
while(x.vals[i] > 0) {
  r <- runif(1, min = -2, max = 1)
  x.vals <- c(x.vals, x.vals[i] + r)
  i <- i+1
}
x.vals
```

(14) [5 points] Produce a plot of the random walk values `x.vals` from above versus the iteration number. Make sure the plot has an appropriately labeled x-axis and and y-axis. Also use `type="o"` so that we can see both points and lines.
```{r}
plot(1:length(x.vals), x.vals, type = "o", xlab = "Iteration", ylab = "Random Walk")
```

(15) [5 points] Write a function `random.walk()` to perform the random walk procedure that you implemented in question (9). Its inputs should be: `x.start`, a numeric value at which we will start the random walk, which takes a default value of $5$; and `plot.walk`, a boolean value, indicating whether or not we want to produce a plot of the random walk values `x.vals` versus the iteration number as a side effect, which takes a default value of `TRUE`. The output of your function should be a list with elements: `x.vals`, a vector of the random walk values as computed above; and `num.steps`, the number of steps taken by the random walk before terminating. Run your function twice with the default inputs, and then twice times with `x.start` equal to $10$ and `plot.walk = FALSE`.
```{r}
random.walk <- function(x.start = 5, plot.walk = TRUE) {
  x.vals <- 5
  i <- 1
  while(all(x.vals[i] > 0)) {
    r <- runif(1, min = -2, max = 1)
    x.vals <- c(x.vals, x.vals[i] + r)
    i <- i+1
  }
  if (plot.walk) {plot(1:length(x.vals), x.vals, type = "o", xlab = "Iteration", ylab = "Random Walk")}
  return(list(x.vals = x.vals, num.steps = i))
}

random.walk()
random.walk()

random.walk(x.start = 10, plot.walk = FALSE)
random.walk(x.start = 10, plot.walk = FALSE)
```

(16) [5 points] We'd like to answer the following question using simulation: if we start our random walk process, as defined above, at $x=5$, what is the expected number of iterations we need until it terminates?  To estimate the solution produce $10,000$ such random walks and calculate the average number of iterations in the $10,000$ random walks you produce.  You'll want to turn the plot off here.
```{r}
n     <- 10000
iters <- rep(NA, n)
for (i in 1:n) {
  iters[i] <- random.walk(plot.walk = FALSE)$num.steps
}
mean(iters)
```
  
(17) [5 points] Modify your function `random.walk()` defined previously so that it takes an additional argument `seed`: this is an integer that should be used to set the seed of the random number generator, before the random walk begins, with `set.seed()`. But, if seed is `NULL`, the default, then no seed should be set. Run your modified function `random.walk()` function several times with the default inputs, then run it several times with the input seed equal to (say) $33$. 

```{r}
random.walk <- function(x.start = 5, plot.walk = TRUE, seed = NULL) {
  if (!is.null(seed)) {set.seed(seed)}
  x.vals <- 5
  i <- 1
  while(all(x.vals[i] > 0)) {
    r <- runif(1, min = -2, max = 1)
    x.vals <- c(x.vals, x.vals[i] + r)
    i <- i+1
  }
  if (plot.walk) {plot(1:length(x.vals), x.vals, type = "o", xlab = "Iteration", ylab = "Random Walk")}
  return(list(x.vals = x.vals, num.steps = i))
}

random.walk()
random.walk()

random.walk(seed = 33, plot.walk = FALSE)
random.walk(seed = 33, plot.walk = FALSE)
```

# Part 4: Monte Carlo Integration


The goal of this exercise is to estimate the mathematical constant $\pi$ using **Monte Carlo Integration**.  Consider the function
\begin{equation}\label{e:qcircle}
g(x)=\begin{cases} \sqrt{1-x^2} \ \ \ \ 0 \leq x \leq 1 \\  0 \ \ \ \   \text{otherwise}
\end{cases}
\end{equation} 
The above function traces out a quartile circle over the interval $[0,1]$.    

Perform the following tasks: 

18) [5 points] Run the following code:
```{r}
g <- function(x) {
  return(sqrt(1-x^2))
} 
plot(seq(0,1,.01),g(seq(0,1,.01)),type="l",col="purple")
```
The above code should produce the plot of a quarter circle.  

19) [5 points] Identify the true area under the curve $g(x)$ by using simple geometric formulas.   

\[
\frac{\pi r^2}{4} = \frac{\pi*1^2}{4}=\frac{\pi}{4} 
\]


20) [5 points] Using **Monte Carlo Integration**, approximate the mathematical constant $\pi$ within a 1/1000 of the true value.  When performing this simulation, make sure to choose a probability density function that has support over the unit interval, i.e. uniform(0,1) or beta($\alpha$,$\beta$).   

```{r}
# Choose n=1000^2 
X <- runif(1000^2)
pi.est <- mean(4*sqrt(1-X^2))
pi.est
abs(pi-pi.est)
```





\large{ \textbf{Please submit the knitted .pdf file!}}
