---
title: "hw7_yw3204"
author: "wyh"
date: "11/25/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## i.
```{r}
poisLoglik <- function(l, data) {
  num <- (l ** data) * exp(-l)
  deno <- factorial(data)
  ratio <- num / deno
  res <- sum(log(ratio))
  return(res)
}

poisLoglik(1, c(1, 0, 0, 1, 1))
```

## ii.
```{r}
# read data
moretti <- read.csv("moretti.csv")

count_new_genres <- function(year) {
  return(sum(moretti$Begin == year))
}

count_new_genres(1803)
count_new_genres(1850)
```
The values are 0 and 3 seperately.

## iii.
```{r}
new_genres <- c()

for(i in c(1740:1900)) {
  new_genres <- c(new_genres, count_new_genres(i))
}

new_genres[64]
new_genres[111]
```
The positions in the vector correspond to the years 1803 and 1850 are 64 and 111. The values should be 0 and 3 which are exactly what in the corresponding positions in vector new_genres. 

## iv.
```{r}
lambdas <- seq(0.01, 1, 0.01)
logs <- c()
for(l in lambdas) {
  logs <- c(logs, poisLoglik(l, new_genres))
}
plot(lambdas, logs, type = "l", xlab = "lambda", ylab = "log-likelihood")
```
The maximum is reached at around 0.2.

## v.
```{r}
negPoisLoglik <- function(l, data = new_genres) {
  num <- (l ** data) * exp(-l)
  deno <- factorial(data)
  ratio <- num / deno
  res <- sum(log(ratio))
  return(-res)
}

nlm(negPoisLoglik, p = 1)
```
According to the result above, the minimum of the negative log-likelihhod function is reached at around 0.273. Thus, the maximum of the log-likelihhod function is reached at around 0.273.

## vi.
```{r}
# store the time interval
year_lap <- c()

# find those year which has new genres
inds <- which(new_genres != 0)

year_lap <- diff(inds)
# for(i in c(1:length(inds)-1)) {
#   year_lap <- c(year_lap, inds[i+1] - inds[i])
# }

year_lap
mean(year_lap)
sd(year_lap)
coe_var <- sd(year_lap)/mean(year_lap)
```

## vii.
### a.
```{r}
cal_interval <- function(vec) {
  inds <- which(vec != 0)
  intergenre_intervals <- diff(inds)
  return(intergenre_intervals)
}

cal_interval(new_genres)
```

### b.
```{r}
cal_CoeVar <- function(num_ys, lambda) {
  n_genres <- rpois(num_ys, lambda)
  
  intervals <- cal_interval(n_genres)
  
  coe_var <- sd(intervals)/mean(intervals)
  
  return(list(intervals, coe_var))
}

cal_CoeVar(161, 0.273)

# chek teh mean
# mean(cal_CoeVar(162, 0.273)[[1]])
```

## viii.
```{r}
coes <- c()

for(i in c(1:10000)) {
  coes <- c(coes, cal_CoeVar(161, 0.273)[[2]])
}

# fraction of simulations having a higher coefficient of variation than Moretti’s data
sum(coes > coe_var) / length(coes)
```

## ix.
The proportion is relatively small and thus tells us that we conjecture that genres don't tend to appear together in burst.







