---
title: "Homework 7 Solutions Help"
output: pdf_document
---



# Part 1

```{r}
genres <- read.csv("moretti.csv", as.is = TRUE)
head(genres)
```

i.

```{r}
poisLoglik <- function(lambda, data) {
  return(sum(dpois(data, lambda = lambda, log = TRUE)))
}

data1 <- c(1, 0, 0, 1, 1)
lambda1 <- 1

poisLoglik(lambda1, data1)
poisLoglik(0, data1)
```

ii.

```{r}
count_new_genres <- function(year) {
  return(sum(genres$Begin == year))
}


count_new_genres(1803)
count_new_genres(1850)

genres$Begin
which(genres$Begin==1846)
#genres
```


iii.

```{r}
years <- seq(1740, 1900)
num.years <- length(years)
new_genres <- rep(NA, num.years)
names(new_genres) <- years

for (i in 1:num.years) {
  new_genres[i] <- count_new_genres(years[i])
}

twoyears <- which(years == 1803 | years == 1850)
twoyears
new_genres[twoyears]
new_genres
```


iv.
```{r}
lambda_seq  <- seq(0, 3, length.out = 100)
num.lambdas <- length(lambda_seq)
Logliks    <- rep(NA, num.lambdas)
for (i in 1:num.lambdas) {
  Logliks[i] <- poisLoglik(lambda_seq[i], new_genres)
}
library(ggplot2)
ggplot() + 
  geom_line(mapping = aes(x = lambda_seq, y = Logliks)) +
  labs(main = "Plot of the Log Likelihood Function", ylab = "Log Likelihood", xlab = "Lambda")

lambda_seq[which.max(Logliks)]
```

v.
```{r}
NegpoisLoglik <- function(lambda, data) {
  return(-sum(dpois(data, lambda = lambda, log = TRUE)))
}
nlm(NegpoisLoglik, 1, new_genres)
```

vi.

```{r}
#genres$Begin
intergenre_intervals <- diff(sort(genres$Begin))
intergenre_intervals 
mean(intergenre_intervals)
sd(intergenre_intervals)
moretti.coef <- sd(intergenre_intervals)/mean(intergenre_intervals)
```

vii. 

a.
```{r}
# Look at body of the function 
names(new_genres) <- 1:length(new_genres)
new_genres
new_genres        <- new_genres[new_genres !=0]
new_genres
 years             <- as.numeric(rep(names(new_genres), new_genres))
years

# define function
intergenre_calc <- function(new.genres) {
  names(new.genres) <- 1:length(new.genres)
  new.genres        <- new.genres[new.genres !=0]
  years             <- as.numeric(rep(names(new.genres), new.genres))
  return(diff(sort(years)))
}


intergenre_intervals2 <- intergenre_calc(new_genres)
all(intergenre_intervals == intergenre_intervals2)
```

b.
```{r}
Pois.sim <- function(num.years, mean.genres) {
  samples <- rpois(num.years, lambda = mean.genres)
  intergenre_intervals <- intergenre_calc(samples)
  coef.of.var <- sd(intergenre_intervals)/mean(intergenre_intervals)
  return(list(intergenre_intervals = intergenre_intervals, coef.of.var = coef.of.var))
}

for (i in 1:10) {
  res <- Pois.sim(161, 0.273)
  print(mean(res$intergenre_intervals))
}
```

viii.
```{r}
n <- 10000
coef.of.var <- rep(NA, n)

for (i in 1:n) {
  res <- Pois.sim(161, 0.273)
  coef.of.var[i] <- res$coef.of.var
}

mean(coef.of.var > moretti.coef)
```

ix.

The above result tells us that there isn't really any evidence that new genres appear together in bursts.  If the appearance of new genres were truly random (not clustered), meaning they appear according to a Poisson process, then around 23 percent of the time we would see results as or more clustered than Moretti's data (when we consider the coefficient of variation as a measure of the amount of cluster).

